Approximation as Differentiation

### 2025-04-15 appendix/notes.tex
Move notes to appendix.
mathcal for metavariables that range over categories.

\documentclass[acmsmall,screen]{acmart}

\usepackage{bbm}
\usepackage{tikz-cd}

\geometry{paperwidth=8.3in, paperheight=11.7in} % force to A4 for now
\settopmatter{printacmref=false}

\input{macros}

\begin{document}

\title{Approximation as Differentiation notes}

\author{Robert Atkey}
\email{robert.atkey@strath.ac.uk}
\orcid{0000-0002-4414-5047}
\affiliation{%
  \institution{University of Strathclyde}
  \city{Glasgow}
  \country{UK}}

\author{Roly Perera}
\email{roly.perera@cl.cam.ac.uk}
\orcid{0000-0001-9249-9862}
\affiliation{%
  \institution{University of Cambridge}
  \city{Cambridge}
  \country{UK}
}
\additionalaffiliation{%
   \institution{University of Bristol}
   \city{Bristol}
   \country{UK}
}

\maketitle

\section{Overview}

Covered so far:
\begin{itemize}
\item $\CMon$-enriched categories, including enriched functors and enriched Yoneda embedding (\secref{cmon-enriched})
\item biproducts and semi-additive categories (\secref{biproduct})
\item category $\Fam(I,C)$ of $I$-indexed families of objects of $C$ (\secref{fam})
\item set-indexed products (\secref{set-indexed-product})
\item Grothendieck construction $\Grothendieck{C}F$ and category of families $\Fam(C)$ (\secref{grothendieck})
\item Examples of useful semi-additive categories (\secref{useful-semi-additive-categories}):
   \begin{itemize}
   \item category $\LatGal$ of bounded lattices and Galois connections (\secref{categories-with-biproducts:latgal})
   \item category $\FinVect_F$ of finite-dimensional vector spaces over a field $F$
   (\secref{categories-with-biproducts:fdvect})
   \end{itemize}
\end{itemize}

@@

\input{notes/sec/notation}

\section{Definitions}

\input{notes/sec/cmon-enriched}
\input{notes/sec/biproduct}
\input{notes/sec/fam}
\input{notes/sec/set-indexed-product}
\input{notes/sec/grothendieck}
\input{notes/sec/useful-semi-additive-categories}

% \input{notes/language}

\section{Some references}

\cite{karvonen2020}

\bibliographystyle{plainnat}
\bibliography{bib}

\end{document}

>>>>

\section{Notes}

\subsection{Overview}

Covered so far:
\begin{itemize}
\item $\CMon$-enriched categories, including enriched functors and enriched Yoneda embedding (\secref{cmon-enriched})
\item biproducts and semi-additive categories (\secref{biproduct})
\item category $\Fam(I,\cat{C})$ of $I$-indexed families of objects of $\cat{C}$ (\secref{fam})
\item set-indexed products (\secref{set-indexed-product})
\item Grothendieck construction $\Grothendieck{\cat{C}}F$ and category of families $\Fam(\cat{C})$
(\secref{grothendieck})
\item Examples of useful semi-additive categories (\secref{useful-semi-additive-categories}):
   \begin{itemize}
   \item category $\LatGal$ of bounded lattices and Galois connections (\secref{categories-with-biproducts:latgal})
   \item category $\FinVect_F$ of finite-dimensional vector spaces over a field $F$
   (\secref{categories-with-biproducts:fdvect})
   \end{itemize}
\end{itemize}

@@

\input{appendix/notes/notation}
\input{appendix/notes/cmon-enriched}
\input{appendix/notes/biproduct}
\input{appendix/notes/fam}
\input{appendix/notes/set-indexed-product}
\input{appendix/notes/grothendieck}
\input{appendix/notes/useful-semi-additive-categories}

% \input{appendix/notes/language}

### 2025-04-15 appendix/notes/cmon-enriched.tex
mathcal for metavariables that range over categories.

Recall that if a category $C$ is $\CMon$-enriched, then:
\begin{enumerate}
\item Every hom-object $\Hom{C}{A}{B}$ is a commutative monoid of morphisms; we write $\zero_{A,B}$ (zero
morphism) for the unit, and $+_{A,B}$ (addition of morphisms) for the binary operation.
\item Composition is \emph{bilinear}, i.e.~given by a family of morphisms $\Hom{C}{B}{C} \tensor
\Hom{C}{A}{B} \to \Hom{C}{A}{C}$ in $\CMon$ that preserve the additive structure in $\Hom{C}{B}{C}$ and
$\Hom{C}{A}{B}$ separately:

@@

\begin{proposition}
If $C$ is $\CMon$-enriched then any functor category $\Func{D}{C}$ is $\CMon$-enriched.
\end{proposition}

@@

\begin{definition}[$\mathscr{V}$-enriched functor]
For any monoidal category $(\mathscr{V}, \tensor, I)$ and $\mathscr{V}$-enriched categories $C$ and $D$, a
\emph{$\mathscr{V}$-enriched} functor $F: C \to D$ on morphisms is a family of morphisms $F_{X,Y}:
\Hom{C}{X}{Y} \to \Hom{D}{F(X)}{F(Y)}$ between hom-objects in $\mathscr{V}$ that makes the following diagrams
in $\mathscr{V}$ commute:

\begin{center}
\begin{tikzcd}[column sep=1.8cm]
I \arrow[d, equals] \arrow[r, "\id_X"] & \Hom{C}{X}{X} \arrow[d, "F_{X,X}"] \\
I \arrow[r, "\id_{F(X)}"'] & \Hom{D}{F(X)}{F(X)}
\end{tikzcd}
\hspace{5mm}
\begin{tikzcd}[column sep=2.6cm]
   \Hom{C}{Y}{Z} \tensor \Hom{C}{X}{Y} \arrow[r, "\comp_{X,Y,Z}"] \arrow[d, "F_{Y,Z} \tensor F_{X,Y}"'] & \Hom{C}{X}{Z} \arrow[d, "F_{X,Z}"] \\
   \Hom{D}{F(Y)}{F(Z)} \tensor \Hom{D}{F(X)}{F(Y)} \arrow[r, "\comp_{F(X),F(Y),F(Z)}"'] & \Hom{D}{F(X)}{F(Z)}
\end{tikzcd}
\end{center}
\end{definition}

Write $\Func{C}{D}_{\mathscr{V}}$ for the category of $\mathscr{V}$-enriched functors from $C$ to $D$.

In the specific setting of a $\CMon$-enriched presheaf $F: C^\op \to \CMon$, enrichment means that $F$
preserves the additive structure of morphisms:
\begin{itemize}
\item $F(0_{X,Y}) = 0_{F(Y),F(X)}$
\item $F(f + g) = F(f) + F(g)$
\end{itemize}
It then follows automatically that $F$ also preserves bilinear composition, i.e.:
\begin{itemize}
\item $F(f) \comp F(0) = F(f) = F(0) \comp F(f)$
\item $F(f + g) \comp F(h) = F(h \comp f) + F(h \comp g)$
\item $F(h) \comp F(f + g) = F(f \comp h) + F(g \comp h)$
\end{itemize}
in addition to the usual enriched functor properties.

@@

Suppose $C$ a small category. The usual \emph{Yoneda embedding} for $C$ is the functor $\Yoneda: C \to
\Func{C^\op}{\Set}$ sending:
\begin{itemize}
\item any object $X$ in $C$ to the contravariant hom-functor $\Yoneda(X) = \Hom{C}{-}{X}: C^\op \to \Set$
which sends:
   \begin{itemize}
   \item any object $Y$ in $C$ to the hom-set $\Hom{C}{Y}{X}$;
   \item any morphism $f: Y \to Z$ in $C$ to the function $\Hom{C}{f}{X} = (- \comp f): \Hom{C}{Z}{X} \to
   \Hom{C}{Y}{X}$.
   \end{itemize}
\item any morphism $f: X \to Y$ in $C$ to the natural transformation $\Yoneda(f)$ where $\Yoneda(f)_Z = (f
\comp -): \Hom{C}{Z}{X} \to \Hom{C}{Z}{Y}$.
\end{itemize}

\noindent Now suppose $C$ is $\mathscr{V}$-enriched with $\mathscr{V}$ closed monoidal; then the hom-functor
$\Hom{C}{-}{X}: C^\op \to \mathscr{V}$ is $\mathscr{V}$-enriched. The following diagram commutes because
precomposition with the identity is the identity on morphisms:

\begin{center}
\begin{tikzcd}[column sep=1.8cm]
I \arrow[d, equals] \arrow[r, "\id_X"] & \Hom{C}{X}{X} \arrow[d, "\Hom{C}{-}{W}_{X,X}"] \\
I \arrow[r, "\id_{\Hom{C}{X}{W}}"'] & \Hom{\mathscr{V}}{\Hom{C}{X}{W}}{\Hom{C}{X}{W}}
\end{tikzcd}
\end{center}

@@

\begin{center}
\begin{tikzcd}[column sep=2.7cm]
   \Hom{C}{Z}{Y} \tensor \Hom{C}{Y}{X} \arrow[r, "\comp_{X,Y,Z}"] \arrow[d, "\Hom{C}{-}{W}_{Y,Z}\,\tensor\,\Hom{C}{-}{W}_{X,Y}"'] & \Hom{C}{Z}{X} \arrow[d, "\Hom{C}{-}{W}_{X,Z}"] \\
   \Hom{\mathscr{V}}{\Hom{C}{Y}{W}}{\Hom{C}{Z}{W}} \tensor \Hom{\mathscr{V}}{\Hom{C}{X}{W}}{\Hom{C}{Y}{W}} \arrow[r, "\comp_{\Hom{C}{X}{W},\Hom{C}{Y}{W},\Hom{C}{Z}{W}}"'] & \Hom{\mathscr{V}}{\Hom{C}{X}{W}}{\Hom{C}{Z}{W}}
\end{tikzcd}
\end{center}

\noindent \todo{Then there is an enriched version of the Yoneda embedding, namely a functor
$\Yoneda_{\mathscr{V}}: C \to \Func{C^\op}{\mathscr{V}}_{\mathscr{V}}$ analogous to $\Yoneda$ which is itself
$\mathscr{V}$-enriched.}

>>>>

Recall that if a category $\cat{C}$ is $\CMon$-enriched, then:
\begin{enumerate}
\item Every hom-object $\Hom{\cat{C}}{A}{B}$ is a commutative monoid of morphisms; we write $\zero_{A,B}$
(zero morphism) for the unit, and $+_{A,B}$ (addition of morphisms) for the binary operation.
\item Composition is \emph{bilinear}, i.e.~given by a family of morphisms $\Hom{\cat{C}}{B}{C} \tensor
\Hom{\cat{C}}{A}{B} \to \Hom{\cat{C}}{A}{C}$ in $\CMon$ that preserve the additive structure in
$\Hom{\cat{C}}{B}{C}$ and $\Hom{\cat{C}}{A}{B}$ separately:

@@

\begin{proposition}
If $C$ is $\CMon$-enriched then any functor category $\Func{\cat{D}}{\cat{C}}$ is $\CMon$-enriched.
\end{proposition}

@@

\begin{definition}[$\mathscr{V}$-enriched functor]
For any monoidal category $(\mathscr{V}, \tensor, I)$ and $\mathscr{V}$-enriched categories $\cat{C}$ and
$\cat{D}$, a \emph{$\mathscr{V}$-enriched} functor $F: \cat{C} \to \cat{D}$ on morphisms is a family of
morphisms $F_{X,Y}: \Hom{\cat{C}}{X}{Y} \to \Hom{\cat{D}}{F(X)}{F(Y)}$ between hom-objects in $\mathscr{V}$
that makes the following diagrams in $\mathscr{V}$ commute:

\begin{center}
\begin{tikzcd}[column sep=1.8cm]
I \arrow[d, equals] \arrow[r, "\id_X"] & \Hom{C}{X}{X} \arrow[d, "F_{X,X}"] \\
I \arrow[r, "\id_{F(X)}"'] & \Hom{D}{F(X)}{F(X)}
\end{tikzcd}
\hspace{5mm}
\begin{tikzcd}[column sep=2.6cm]
   \Hom{\cat{C}}{Y}{Z} \tensor \Hom{\cat{C}}{X}{Y} \arrow[r, "\comp_{X,Y,Z}"] \arrow[d, "F_{Y,Z} \tensor F_{X,Y}"']
   & \Hom{\cat{C}}{X}{Z} \arrow[d, "F_{X,Z}"] \\
   \Hom{\cat{D}}{F(Y)}{F(Z)} \tensor \Hom{\cat{D}}{F(X)}{F(Y)} \arrow[r, "\comp_{F(X),F(Y),F(Z)}"'] &
   \Hom{\cat{D}}{F(X)}{F(Z)}
\end{tikzcd}
\end{center}
\end{definition}

\begin{definition}[Category of enriched functors]
Write $\Func{\cat{C}}{\cat{D}}_{\mathscr{V}}$ for the category of $\mathscr{V}$-enriched functors from
$\cat{C}$ to $\cat{D}$.
\end{definition}

In the specific setting of a $\CMon$-enriched presheaf $F: \cat{C}^\op \to \CMon$, enrichment means that $F$
preserves the additive structure of morphisms:
\begin{itemize}
\item $F(0_{X,Y}) = 0_{F(Y),F(X)}$
\item $F(f + g) = F(f) + F(g)$
\end{itemize}
It then follows automatically that $F$ also preserves bilinear composition, i.e.:
\begin{itemize}
\item $F(f) \comp F(0) = F(f) = F(0) \comp F(f)$
\item $F(f + g) \comp F(h) = F(h \comp f) + F(h \comp g)$
\item $F(h) \comp F(f + g) = F(f \comp h) + F(g \comp h)$
\end{itemize}
in addition to the usual enriched functor properties.

@@

Suppose $\cat{C}$ a small category. The usual \emph{Yoneda embedding} for $\cat{C}$ is the functor $\Yoneda:
\cat{C} \to \Func{\cat{C}^\op}{\Set}$ sending:
\begin{itemize}
\item any object $X$ in $\cat{C}$ to the contravariant hom-functor $\Yoneda(X) = \Hom{\cat{C}}{-}{X}:
\cat{C}^\op \to \Set$
which sends:
   \begin{itemize}
   \item any object $Y$ in $\cat{C}$ to the hom-set $\Hom{\cat{C}}{Y}{X}$;
   \item any morphism $f: Y \to Z$ in $\cat{C}$ to the function $\Hom{\cat{C}}{f}{X} = (- \comp f):
   \Hom{\cat{C}}{Z}{X} \to \Hom{\cat{C}}{Y}{X}$.
   \end{itemize}
\item any morphism $f: X \to Y$ in $\cat{C}$ to the natural transformation $\Yoneda(f)$ where $\Yoneda(f)_Z =
(f \comp -): \Hom{\cat{C}}{Z}{X} \to \Hom{\cat{C}}{Z}{Y}$.
\end{itemize}

\noindent Now suppose $\cat{C}$ is $\mathscr{V}$-enriched with $\mathscr{V}$ closed monoidal; then the
hom-functor $\Hom{\cat{C}}{-}{X}: \cat{C}^\op \to \mathscr{V}$ is $\mathscr{V}$-enriched. The following
diagram commutes because precomposition with the identity is the identity on morphisms:

\begin{center}
\begin{tikzcd}[column sep=1.8cm]
I \arrow[d, equals] \arrow[r, "\id_X"] & \Hom{\cat{C}}{X}{X} \arrow[d, "\Hom{\cat{C}}{-}{W}_{X,X}"] \\
I \arrow[r, "\id_{\Hom{\cat{C}}{X}{W}}"'] & \Hom{\mathscr{V}}{\Hom{\cat{C}}{X}{W}}{\Hom{\cat{C}}{X}{W}}\
\end{tikzcd}
\end{center}

@@

\begin{center}
\begin{tikzcd}[column sep=2.7cm]
   \Hom{\cat{C}}{Z}{Y} \tensor \Hom{\cat{C}}{Y}{X} \arrow[r, "\comp_{X,Y,Z}"] \arrow[d,
   "\Hom{\cat{C}}{-}{W}_{Y,Z}\,\tensor\,\Hom{\cat{C}}{-}{W}_{X,Y}"'] & \Hom{\cat{C}}{Z}{X} \arrow[d,
   "\Hom{\cat{C}}{-}{W}_{X,Z}"] \\
   \Hom{\mathscr{V}}{\Hom{\cat{C}}{Y}{W}}{\Hom{\cat{C}}{Z}{W}} \tensor
   \Hom{\mathscr{V}}{\Hom{\cat{C}}{X}{W}}{\Hom{\cat{C}}{Y}{W}} \arrow[r,
   "\comp_{\Hom{\cat{C}}{X}{W},\Hom{\cat{C}}{Y}{W},\Hom{\cat{C}}{Z}{W}}"'] &
   \Hom{\mathscr{V}}{\Hom{\cat{C}}{X}{W}}{\Hom{\cat{C}}{Z}{W}}
\end{tikzcd}
\end{center}

\noindent \todo{Then there is an enriched version of the Yoneda embedding, namely a functor
$\Yoneda_{\mathscr{V}}: \cat{C} \to \Func{\cat{C}^\op}{\mathscr{V}}_{\mathscr{V}}$ analogous to $\Yoneda$
which is itself $\mathscr{V}$-enriched.}
